\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-17 paper/main.tex
JRM edits (next pass)

%%
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01

\documentclass[preprint,12pt,a4paper]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

\usepackage{float}
\restylefloat{table}

% ======== additional package not in SoftwareX template ========
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage[htt]{hyphenat}
%\usepackage{listings}
%\usepackage{xcolor}
%\usepackage{todonotes}
\usepackage{placeins}

% \let\oldtodo\todo
% \renewcommand{\todo}[1]{\oldtodo[tickmarkheight=0.5em]{#1}}

\newcommand{\comment}[1]{}
%\let\oldtexttt\texttt
%\renewcommand{\texttt}[1]{\mbox{\oldtexttt{#1}}}
% =================================================

@@

% ------------------------------------------------------ ABSTRACT ------------------------------------------------------

\begin{abstract}
% 100-word limit (though many published articles don't seem to conform
% to this?)

  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the
  complexity and time cost of sharing or engaging with reproducible
  science. Here, we propose a lighter-weight solution:  the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, \texttt{davos} provides a
  mechanism for specifying (and automatically installing) the correct
  versions of the project's dependencies.  The \texttt{davos}
  library further ensures that those packages and specific versions are used every time the
  notebook's code is executed.  This enables researchers to share a complete
  reproducible copy of their code within a single Jupyter notebook file.

\end{abstract}

@@

% ------------------------------------------------------ METADATA ------------------------------------------------------
\section*{Required Metadata}

@@

% --------------------------------------------- MOTIVATION & SIGNIFICANCE ----------------------------------------------
\section{Motivation and significance}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external libraries, but
  those libraries must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those libraries were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} library} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external libraries (at their
  correct versions, when specified).  \textbf{Virtual environments}
  allow users to install an isolated copy of Python and all required
  dependencies. This typically entails distributing a configuration file
  (e.g., a \texttt{pyproject.toml}~\cite{CannEtal16} or \texttt{environment.yml})
  that specifies all project dependencies (including version numbers of
  external libraries) alongside the primary code base. Users can then install
  a third-praty tool~\cite[e.g.,][]{Anac12, Eust19} to read the file and build the environment.
  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to (optionally)
  specifying a virtual environment or other configurations needed to
  provide the necessary computing environment.  Containers are
  typically defined using specification files (e.g., a plain-text
  \texttt{Dockerfile}) that instruct the virtualization engine
  regarding how to build the containerized environment.  \textbf{Virtual
    machines} provide a complete hardware-level simulation of the
  computing environment.  In addition to simulating specific hardware,
  virtual machines (typically specified using binary images files)
  must also define operating system-level properties of the computing
  environment.  Systems to the left of the blue vertical line entail
  sharing individual files, with no additional installation or
  configuration needed to run the target code.  Systems to the right
  of the red vertical line support precise control over dependencies
  and versioning.  Notebooks enhanced using the \texttt{davos} library
  are easily shareable and require minimal setup costs, while also
  facilitating high reproducibility by enabling precise control over
  project dependencies.}
\label{fig:code-sharing}
\end{figure}

Computational researchers and other programmers have de\-vel\-oped a broad
set of approaches and tools to facilitate code sharing and
reproducible outcomes (Fig.~\ref{fig:code-sharing}). At one
extreme, simply distributing a set of Python scripts (\texttt{.py} files) may
enable others to use or gain insights into the relevant work. Because
Python is installed by default on most modern operating systems, for
some projects, this may be sufficient. Another popular approach
entails creating Jupyter
notebooks~\cite{KluyEtal16} that comprise a mix of text, executable
code, and embedded media. Notebooks may call or import external
scripts or libraries---even intersperse snippets of other programming
or markup lang\-uages---in order to provide a more compact and readable
experience for users. Both of these systems (Python scripts and
notebooks) provide a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed. Therefore the functionality of code shared using
these systems cannot be guaranteed across different users or setups.

At another extreme, virtual machines~\cite{Gold74, AltiEtal05,
 Rose99} provide a hardware-level simulation of the desired system.
Virtual machines are typically isolated such
that installing or running software on a virtual machine does not
impact the user's primary operating system or computing environment.
Containers~\cite[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience. Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies. Virtual
environments~\cite[e.g.,][]{Anac12, Eust19} also provide a computing
environment that is largely separated from the user's main
environment. They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment. Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees---at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users. However, each
of these systems also relies on additional software that can be
complex or resource-intensive to install and use, creating potential
barriers to both contributing to and taking advantage of open science
resources.

@@

% ------------------------------------------------ SOFTWARE DESCRIPTION ------------------------------------------------
\section{Software description}
% Describe the software in as much as is necessary to establish a vocabulary needed to explain its impact.
The \texttt{davos} package is named after Davos Seaworth, a smuggler often referred to as ``the Onion Knight" from the series \textit{A Song of Ice and Fire} by George R. R. Martin.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} }
\label{fig:package-structure}
\end{figure}

\subsection{Software architecture}
% Give a short overview of the overall software architecture; provide a pictorial component overview or similar (if possible). If necessary provide implementation details.
The \texttt{davos} package consists of two interdependent subpackages (see Fig.~\ref{fig:package-structure}). The first, \texttt{davos.core}, comprises a set of modules that implement\comment{provide?} the bulk of the package's core functionality, including pipelines for installing and validating packages, custom parsers for the \texttt{smuggle} statement (see Section~\ref{subsec:smuggle}) and onion comment (see Section~\ref{subsec:onion}), and a runtime interface for configuring \texttt{davos}'s behavior (see Section~\ref{subsec:config}). However, certain critical\comment{other important} aspects of this functionality require (often substantially) different implementations depending on properties of the notebook environment in which \texttt{davos} is used (e.g., whether the frontend is provided by Jupyter or Google Colaboratory, or which version of IPython~\cite{PereGran07} is used by the notebook kernel). To deal with this, environment-dependent parts of core features and behaviors are isolated and abstracted to ``helper functions'' in the \texttt{davos.implementations} subpackage. This second subpackage defines multiple, interchangeable versions of each helper function, organized into modules by the conditions that trigger their use. At runtime, \texttt{davos} detects various features in the notebook environment and selectively imports a single version of each helper function into the top-level \texttt{davos.implementations} namespace, allowing \texttt{davos.core} modules to access the proper implementations for the current notebook environment in a single, consistent location. An additional benefit of this design pattern is that it allows both maintainers and users to easily extend \texttt{davos} to support new, updated, or custom notebook variants by creating a new \texttt{davos.implementations} module that defines its own version of each helper function, modified as much or little as necessary.
%The package also includes stub files (\texttt{.pyi} files) for each module for use with a static type checker, as well as a custom GitHub Actions-based CI test suite that uses

\comment{
- js_functions.py?
- how parser is registered and deregistered?
- stub files?
- test suite?
- packaged with new PEP \_\_\_ standard (pyproject.toml + setup.cfg; no setup.py)?
}

\subsection{Software functionalities}
% Present the major functionalities of the software.
\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}
Functionally, importing \texttt{davos} in an IPython notebook enables an additional Python keyword: ``\texttt{smuggle}'' (see Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} statement can be used as a drop-in replacement for Python's built-in ``\texttt{import}'' to load libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is not installed locally, \texttt{smuggle} statements can handle missing packages on the fly.
If a smuggled package does not exist in the local environment, \texttt{davos} will install it automatically, expose its contents to Python's \texttt{import} machinery, and load it into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}
For greater control over the behavior of \texttt{smuggle} statements, \texttt{davos} defines an additional construct called the ``onion comment.'' An onion comment is a special type of inline comment that may be placed on a line containing a \texttt{smuggle} statement to customize how \texttt{davos} searches for the smuggled package locally and, if necessary, downloads and installs it. Onion comments follow a simple format based on the ``type comment'' syntax introduced in PEP 484~\cite{vanREtal14}, and are designed to make managing packages with \texttt{davos} intuitive and familiar. To construct an onion comment, users provide the name of the installer program (e.g., \texttt{pip}) and the same arguments one would use to manually install the package as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when installing it) may differ from its top-level module name (i.e., the name used when importing it). In such cases, an onion comment can be used to ensure that \texttt{davos} installs the proper distribution if the smuggled package can't be found locally:
%Onion comments are useful in such cases for ensuring \texttt{davos} installs the proper distribution if the smuggled package isn't installed locally:
%Onion comments are useful when smuggling a package whose distribution name (i.e., the name used when installing it) differs from its top-level module name (i.e., the name used when importing it). Onion comments are useful in such cases for
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
However, the most powerful use of the onion comment is making \texttt{smuggle} statements \textit{version-sensitive}. If an onion comment includes a version specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the version of the package loaded into the notebook matches the specific version requested, or satisfies the given version constraints. If the smuggled package exists locally, \texttt{davos} will extract its version info from its metadata and compare it to the specifier provided. If the two are incompatible (or no local installation is found), \texttt{davos} will install and load a suitable version of the package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references (e.g., Git~\cite{TorvHama05}  branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding arguments to the installer program. In addition to preventing onion comments from being used as a vehicle for shell injection attacks, this enables \texttt{davos} to take certain logical actions when particular arguments are passed. For example, each of the \mbox{\texttt{-I}/\texttt{--ignore-installed}}, \texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags will cause \texttt{davos} to skip searching for a smuggled package locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, passing \texttt{--no-input} will temporarily enable \texttt{davos}'s non-interactive mode (see Section~\ref{subsec:config}), and installing a smuggled package into \texttt{<dir>} with \texttt{--target <dir>} will cause \texttt{davos} to prepend \texttt{dir} to the module search path (i.e., \texttt{sys.path}), if necessary, so the package can be imported.

\subsubsection{The \texttt{davos} config}\label{subsec:config}
The \texttt{davos} config provides a simple, high-level interface for controlling various aspects of \texttt{davos}'s behavior. After importing \texttt{davos}, the \texttt{davos.config} object (a singleton) exposes a number of configurable options as attributes that can be assigned different values, checked at runtime, and displayed in the notebook cell output (see Fig.~\ref{fig:illustrative-example} for example usage). These include:
%The \texttt{davos.config} object provides a simple, high-level interface for controlling various aspects of \texttt{davos}'s behavior. After importing \texttt{davos}, the config instance (a singleton) for the current session exposes a number of options that can be set by assigning values to the object's attributes (see Fig.~\textcolor{red}{INSERT REF TO ILLUSTRATIVE EXAMPLE FIG} for example usage). These include:
%The \texttt{davos} config object provides a simple, high-level interface that allows users to view and set various options that affect \texttt{davos}'s behavior. After importing \texttt{davos}, the config instance (a singleton) for the current session is available as \texttt{davos.config}, and its various fields are accessible as attributes (see Fig.~\textcolor{red}{REF TO ILLUSTRATIVE EXAMPLE FIG} for example usage). The config object exposes a mixture of writable and read-only fields. Writable fields include:
\begin{itemize}
\item \texttt{.active}: This option allows users to disable \texttt{davos} functionality (i.e., support for \texttt{smuggle} statements and onion comments) for subsequent notebook cells by setting its value to \texttt{False}. \texttt{davos} can be re-enabled at any time by setting this option to \texttt{True} (the default when \texttt{davos} is first imported). See Section~\ref{subsec:implementation} for additional info.
\item \texttt{.auto\_rerun}: This option controls how \texttt{davos} behaves when attempting to \texttt{smuggle} a new version of a package that was previously imported and cannot be reloaded. This can happen if the package includes extension modules that dynamically link C or C++ objects to the Python interpreter itself, and the code that generates those objects was changed between the old and new versions. If this option is set to \texttt{True}, \texttt{davos} will automatically restart the notebook kernel and rerun all code up to (and including) the current \texttt{smuggle} statement. If \texttt{False} (the deafult), \texttt{davos} will instead issue a warning, pause execution, and prompt the user with buttons to either restart and rerun the notebook, or continue running with the previously imported package version until the next time the kernel is restarted manually. (Note: not configurable in Google Colaboratory).
%\item \texttt{.auto\_rerun}: Some packages (e.g., \texttt{numpy}, \texttt{pandas}) contain extension modules that dynamically link C or C++ objects to the Python interpreter when imported. If \texttt{davos} is used to \texttt{smuggle} a specific version of such a package after a different version of the same package was previously imported
%\item \texttt{.auto\_rerun}: This option controls how \texttt{davos} behaves when attempting to \texttt{smuggle} a new version of a package that was previously imported (at a different version) and cannot be reloaded (i.e., because it contains C-extensions that dynamically generate code). If \texttt{True} (default: \texttt{False}), \texttt{davos} will automatically restart the notebook kernel and rerun all code up to (and including) the current \texttt{smuggle} statement. Otherwise, \texttt{davos} will issue a warning, pause execution, and prompt the user with buttons to either restart and rerun the notebook or continue running with the imported package version. (Note: not configurable in Google Colaboratory).
\item \texttt{.confirm\_install}: If \texttt{True} (default: \texttt{False}), \texttt{davos} will require user confirmation (\texttt{[y]es/[n]o} input) before installing a smuggled package.
\item \texttt{.noninteractive}: Setting this option to \texttt{True} (default: \texttt{False}) enables non-interactive mode, in which all user input and confirmation is disabled. Note that in non-interactive mode, the \texttt{confirm\_install} option is set to \texttt{False}, and if \texttt{auto\_rerun} is \texttt{False}, \texttt{davos} will throw an error if a smuggled package cannot be reloaded, rather than prompting the user.
\item \texttt{.pip\_executable}: This option's value specifies the path to the \texttt{pip} executable used to install smuggled packages. The default is programmatically determined from the Python environment and falls back to \texttt{sys.executable -m pip} if no executable can be found.
\item \texttt{.suppress\_stdout}: If \texttt{True} (default: \texttt{False}), suppress all unnecessary output issued by both \texttt{davos} and the installer program. This can be useful when smuggling packages that need to install many dependencies and/or generate extensive output. If the installer program throws an error, both the stdout and stderr streams will be displayed with the traceback.
\end{itemize}

%\noindent The \texttt{davos.config} object also provides a few informative read-only fields, including:
%\begin{itemize}
%\item \texttt{.environment}: This field's value is a short string describing the notebook environment in which \texttt{davos} is currently running. This is used internally may be one of \texttt{IPython<7.0} (a notebook whose frontend uses the Jupyter JavaScript library and whose backend uses an IPython version older than v7.0), \texttt{IPython>=7.0} (, or \texttt{}
%\end{itemize}
\noindent The top-level \texttt{davos} namespace also defines a handful of convenience functions for setting and checking \texttt{davos}'s active/inactive state (\texttt{davos.activate()}; \texttt{davos.deactivate()}; \texttt{davos.is\_active()}) as well as the \texttt{davos.configure()} function, which allows setting multiple config options at once.

>>>>

\documentclass[preprint,12pt,a4paper]{elsarticle}
\usepackage{amssymb}
\usepackage{lineno}
\usepackage{float}
\restylefloat{table}
\usepackage[colorlinks=true, urlcolor=blue, pdfborder={0 0 0}]{hyperref}
\usepackage{breakurl}
\usepackage[htt]{hyphenat}
\usepackage{placeins}
\newcommand{\comment}[1]{}

@@

\begin{abstract}
  Reproducibility is a core requirement of modern scientific research.
  For computational research, reproducibility means that code should
  produce the same results, even when run on different systems.  A
  standard approach to ensuring reproducibility entails packaging a
  project's dependencies along with its primary code base.  Existing
  solutions vary in how deeply these dependencies are specified,
  ranging from virtual environments, to containers, to virtual
  machines.  Each of these existing solutions requires installing or
  setting up a system for running the desired code, increasing the
  complexity and time cost of sharing or engaging with reproducible
  science. Here, we propose a lighter-weight solution: the
  \texttt{davos} library.  When used in combination with a
  notebook-based Python project, \texttt{davos} provides a mechanism
  for specifying (and automatically installing) the correct versions
  of the project's dependencies.  The \texttt{davos} library further
  ensures that those packages and specific versions are used every
  time the notebook's code is executed.  This enables researchers to
  share a complete reproducible copy of their code within a single
  Jupyter notebook file.
\end{abstract}

@@

\section*{Metadata}

@@

\section{Motivation and significance}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external libraries, but
  those libraries must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those libraries were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} library} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install any required external libraries (at their
  correct versions, when specified).  \textbf{Virtual environments}
  allow users to install an isolated copy of Python and all required
  dependencies. This typically entails distributing a configuration
  file (e.g., a \texttt{pyproject.toml}~\cite{CannEtal16} or
  \texttt{environment.yml} file) that specifies all project
  dependencies (including version numbers of external libraries)
  alongside the primary code base. Users can then install a
  third-party tool~\cite[e.g.,][]{Anac12, Eust19} to read the file and
  build the environment.  \textbf{Containers} provide a means of
  defining an isolated environment that includes a complete operating
  system (independent of the user's operating system), in addition to
  (optionally) specifying a virtual environment or other
  configurations needed to provide the necessary computing
  environment.  Containers are typically defined using specification
  files (e.g., a plain-text \texttt{Dockerfile}) that instruct the
  virtualization engine regarding how to build the containerized
  environment.  \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary image files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing individual files,
  with no additional installation or configuration needed to run the
  target code.  Systems to the right of the red vertical line support
  precise control over dependencies and versioning.  Notebooks
  enhanced using the \texttt{davos} library are easily shareable and
  require minimal setup costs, while also facilitating high
  reproducibility by enabling precise control over project
  dependencies.}
\label{fig:code-sharing}
\end{figure}

Computational researchers and other programmers have de\-vel\-oped a
broad set of approaches and tools to facilitate code sharing and
reproducible outcomes (Fig.~\ref{fig:code-sharing}). At one extreme,
simply distributing a set of Python scripts (\texttt{.py} files) may
enable others to use or gain insights into the relevant work. Because
Python is installed by default on most modern operating systems, for
some projects, this may be sufficient. Another popular approach
entails creating Jupyter notebooks~\cite{KluyEtal16} that comprise a
mix of text, executable code, and embedded media. Notebooks may call
or import external scripts or libraries---or even intersperse snippets
of other programming or markup lang\-uages---in order to provide a
more compact and readable experience for users. Both of these systems
(Python scripts and notebooks) provide a convenient means of sharing
code, with the caveat that they do not specify the computing
environment in which the code is executed. Therefore the functionality
of code shared using these systems cannot be guaranteed across
different users or setups.

At another extreme, virtual machines~\cite{Gold74, AltiEtal05, Rose99}
provide a hardware-level simulation of the desired system.  Virtual
machines are typically isolated, such that installing or running
software on a virtual machine does not impact the user's primary
operating system or computing environment.
Containers~\cite[e.g.,][]{Merk14, KurtEtal17} provide a similar
``isolated'' experience. Although containerized environments do not
specify hardware-level operations, they are typically packaged with a
complete operating system, in addition to a complete copy of Python
and any relevant package dependencies. Virtual
environments~\cite[e.g.,][]{Anac12, Eust19} also provide a computing
environment that is largely separated from the user's main
environment. They incorporate a copy of Python and the target
software's dependencies, but virtual environments do not specify or
reproduce an operating system for the runtime environment. Each of
these systems (virtual machines, containers, and virtual environments)
guarantees (to differing degrees---at the hardware level, operating
system level, and Python environment level, respectively) that the
relevant code will run similarly for different users. However, each of
these systems also relies on additional software that can be complex
or resource-intensive to install and use, creating potential barriers
to both contributing to and taking advantage of open science
resources.

@@

\section{Software description}

The \texttt{davos} package is named after Davos Seaworth, a smuggler
referred to as ``the Onion Knight" from the series \textit{A Song of
  Ice and Fire} by George R. R. Martin~\citep{Mart12}.  The
\texttt{smuggle} keyword-like operator implemented in \texttt{davos}
is a play on Python's \texttt{import} keyword: whereas importing can
bring a package into the Python workspace within the existing rules
and frameworks provided by the Python language, ``smuggling'' provides
an alternative that expands the scope and reach of ``importing.''
Like the character Davos Seaworth (who became famous for smuggling
onions through a blockade on his homeland), we use ``onion'' comments
to precisely control how packages are smuggled into the Python
workspace.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/package_structure}
\caption{\small \textbf{Package structure.} The \texttt{davos} library
  comprises two interdependent subpackages.  The \texttt{davos.core}
  subpackage includes modules for parsing \texttt{smuggle} statements
  and onion comments, installing and validating packages, and
  configuring \texttt{davos}'s behavior.  The
  \texttt{davos.implementations} subpackage includes
  environment-specific modifications and features that are needed to
  support the core functionality across different notebook-based
  environments.  Individual \texttt{.py} files are represented by lime
  rounded rectangles, and arrows denote dependencies (each arrow
  points to a file that import objects defined or specified at the
  arrow's source).}
\label{fig:package-structure}
\end{figure}

\subsection{Software architecture}

The \texttt{davos} package consists of two interdependent subpackages
(see Fig.~\ref{fig:package-structure}). The first,
\texttt{davos.core}, comprises a set of modules that
implement\comment{provide?} the bulk of the package's core
functionality, including pipelines for installing and validating
packages, custom parsers for the \texttt{smuggle} statement (see
Section~\ref{subsec:smuggle}) and onion comment (see
Section~\ref{subsec:onion}), and a runtime interface for configuring
\texttt{davos}'s behavior (see Section~\ref{subsec:config}). However,
certain critical\comment{other important} aspects of this
functionality require (often substantially) different implementations
depending on properties of the notebook environment in which
\texttt{davos} is used (e.g., whether the frontend is provided by
Jupyter or Google Colaboratory, or which version of
IPython~\cite{PereGran07} is used by the notebook kernel). To deal
with this, environment-dependent parts of core features and behaviors
are isolated and abstracted to ``helper functions'' in the
\texttt{davos.implementations} subpackage. This second subpackage
defines multiple, interchangeable versions of each helper function,
organized into modules by the conditions that trigger their use. At
runtime, \texttt{davos} detects various features in the notebook
environment and selectively imports a single version of each helper
function into the top-level \texttt{davos.implementations} namespace,
allowing \texttt{davos.core} modules to access the proper
implementations for the current notebook environment in a single,
consistent location. An additional benefit of this design is that it
allows maintainers, developers, and users to extend \texttt{davos} to
support new, updated, or custom notebook variants by creating new
\texttt{davos.implementations} modules that define their own versions
of each helper function, modified from existing implementations as
needed.

\subsection{Software functionalities}

\subsubsection{The \texttt{smuggle} statement}\label{subsec:smuggle}

Functionally, importing \texttt{davos} in an IPython notebook enables
an additional Python keyword: ``\texttt{smuggle}'' (see
Section~\ref{subsec:implementation} for details on how this works).
The \texttt{smuggle} keyword-like object can be used as a drop-in
replacement for Python's built-in \texttt{import} keyword to load
libraries, modules, and other objects into the current namespace.
However, whereas \texttt{import} will fail if the requested package is
not installed locally, \texttt{smuggle} statements can handle missing
packages on the fly.  If a smuggled package does not exist in the
local environment, \texttt{davos} will install it automatically,
expose its contents to Python's \texttt{import} machinery, and load it
into the namespace for immediate use.

\subsubsection{The onion comment}\label{subsec:onion}

For greater control over the behavior of \texttt{smuggle} statements,
\texttt{davos} defines an additional construct called the ``onion
comment.'' An onion comment is a special type of inline comment that
may be placed on a line containing a \texttt{smuggle} statement to
customize how \texttt{davos} searches for the smuggled package locally
and, if necessary, downloads and installs it. Onion comments follow a
simple format based on the ``type comment'' syntax introduced in PEP
484~\cite{vanREtal14}, and are designed to make managing packages with
\texttt{davos} intuitive and familiar. To construct an onion comment,
users provide the name of the installer program (e.g., \texttt{pip})
and the same arguments one would use to manually install the package
as desired via the command line:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet1}
\end{center}
Occasionally, a package's distribution name (i.e., the name used when
installing it) may differ from its top-level module name (i.e., the
name used when importing it). In such cases, an onion comment can be
used to ensure that \texttt{davos} installs the proper distribution if
the smuggled package cannot be found locally:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet2}
\end{center}
Because onion comments may be constructed to specify any aspect of the
installer's behavior, they provide a mechanism for precisely
controlling how, where, and when smuggled packages are installed.
Critically, if an onion comment includes a version
specifier~\cite{CoghStuf13}, \texttt{davos} will ensure that the
version of the package loaded into the notebook matches the specific
version requested, or satisfies the given version constraints. If the
smuggled package exists locally, \texttt{davos} will extract its
version information from its metadata and compare it to the specifier
provided. If the two are incompatible (or no local installation is
found), \texttt{davos} will install and load a suitable version of the
package instead:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet3}
\end{center}
Onion comments can also be used to smuggle specific VCS references
(e.g., Git~\cite{TorvHama05} branches, commits, tags, etc.):
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet4}
\end{center}
\texttt{davos} processes onion comments internally before forwarding
arguments to the installer program. In addition to preventing onion
comments from being used as a vehicle for shell injection attacks,
this design provides the user with additional control over which
packages are imported. For example, each of the
\mbox{\texttt{-I}/\texttt{--ignore-installed}},
\texttt{-U}/\texttt{--upgrade}, and \texttt{--force-reinstall} flags
will cause \texttt{davos} to skip searching for a smuggled package
locally before installing a new copy:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/snippet5}
\end{center}
Similarly, passing \texttt{--no-input} will temporarily enable
\texttt{davos}'s non-interactive mode (see
Section~\ref{subsec:config}), and installing a smuggled package into a
custom directory (\texttt{<dir>}) using the \texttt{--target <dir>}
flag will cause \texttt{davos} to prepend \texttt{<dir>} to the module
search path (i.e., \texttt{sys.path}), if necessary, so the package
can be imported.

\subsubsection{The \texttt{davos} config object}\label{subsec:config}

The \texttt{davos} config object provides a high-level interface for
controlling various aspects of \texttt{davos}'s behavior. After
importing \texttt{davos}, the \texttt{davos.con\-fig} object (a
singleton) exposes configurable options as attributes that can be
modified, checked at runtime, or displayed (see
Sec.~\ref{sec:illustrative-example} for an illustrative example or
Sec.~\ref{subsec:implementation} for implementation details and
additional information). These include:

\begin{itemize}
\item \texttt{.active}: This attribute controls whether support for \texttt{smuggle}
  statements and onion comments) is enabled (\texttt{True}) or
  disabled (\texttt{False}).  When \texttt{davos} is first imported),
  the \texttt{.active} attribute is set to \texttt{True}.

\item \texttt{.auto\_rerun}: This attribute controls how
  \texttt{davos} behaves when attempting to \texttt{smuggle} a new
  version of a package that was previously imported and cannot be
  reloaded. This can happen if the package includes extension modules
  that dynamically link C or C++ objects to the Python interpreter
  itself, and if the code that generates those objects was changed
  between the previously imported and to-be-smuggled versions.  If
  this attribute is set to \texttt{True}, \texttt{davos} will
  automatically restart the notebook kernel and rerun all code up to
  (and including) the current \texttt{smuggle} statement. If
  \texttt{False} (the default), \texttt{davos} will instead issue a
  warning, pause execution, and prompt the user to either restart and
  rerun the notebook, or continue running with the previously imported
  package version until the next time the kernel is restarted
  manually.  Note that, as of this writing, the \texttt{.auto\_rerun} attribute
  is not supported in Google Colaboratory notebooks.

\item \texttt{.confirm\_install}: If \texttt{True} (default:
  \texttt{False}), \texttt{davos} will require user confirmation
  before installing a smuggled package that do not yet exist in the
  user's environment.

\item \texttt{.noninteractive}: Setting this attribute to
  \texttt{True} (default: \texttt{False}) enables non-interactive
  mode, in which all user interactions (prompts and dialogues) are
  disabled. Note that in non-interactive mode, the
  \texttt{confirm\_install} option is set to \texttt{False}.  If
  \texttt{auto\_rerun} is \texttt{False} while in non-interactive
  mode, \texttt{davos} will raise an exception if a smuggled package
  cannot be reloaded, rather than prompting the user.

\item \texttt{.pip\_executable}: This attribute's value specifies the
  path to the \texttt{pip} executable used to install smuggled
  packages. The default is programmatically determined from the Python
  environment and falls back to \texttt{sys.executable -m pip} if no
  executable can be found.

\item \texttt{.suppress\_stdout}: If this attribute is set to
  \texttt{True} (default: \texttt{False}), \texttt{davos} suppresses
  printed (console) outputs from itself and the installer program.
  This can be useful when smuggling packages that need to install many
  dependencies and/or generate extensive output. However, if the installer
  program throws an error, both the stdout and stderr streams will still be
  displayed along with a stack trace.
\end{itemize}

\noindent The top-level \texttt{davos} namespace also defines
convenience functions for setting and checking whether \texttt{davos}
is active (\texttt{davos.activate()}; \texttt{davos.deactivate()};
\texttt{davos.is\_active()}) as well as the \texttt{davos.configure()}
function, which allows setting multiple configuration options
simultaneously.

\subsection{Implementation details}\label{subsec:implementation}

Although \texttt{davos} is designed to \texttt{appear} to add a new
Python keyword to a notebook's vocabulary, this illusion is actually a
consequence of several ``hacks'' that make use of the IPython backend
for processing and executing notebook code.  Specifically, when
\texttt{davos} is first imported, or when it activated after being set
to an inactive state, two actions are triggered.  First, the
\texttt{smuggle()} function is injected into the IPython user
namespace.  Second, the \texttt{davos} parser is registered as a
custom IPython input transformer.
